# Genetic constitution of a population {#intro}

## Problem 1

The following numbers of the human M-N blood groups were recorded in a sample of American Whites.

|      |      |      |
|------|------|------|
| M    | MN   | N    |
| 1787 | 3039 | 1303 |


**(1) What are the genotype frequencies observed in this sample?**

```{r}
MM <- 1787
MN <- 3039
NN <- 1303
n <- MM + MN + NN

# Genotype frequencies
P <- MM / n
P

H <- MN / n
H

Q <- NN / n
Q
```

(2) What are the gene frequencies?

Given that:

$$p = P + \frac{1}{2}H$$ and $$q = Q + \frac{1}{2}H$$

We get to:

```{r}
# Gene frequencies
p <- P + H/2
p

q <- Q + H/2
```

(3) With the gene frequencies observed, what are the genotype frequencies expected from the Hardy-Weinberg law?

Expected frequencies in the equilibrium:

$$p^2 + 2pq + q^2 = 1$$

```{r}
expected <- c(MM = p^2, 
              MN = 2*p*q, 
              NN = q^2)
expected
```

(4) How well do the observed frequencies agree with the expectation? Data from Wiener, A.S. (1943) quoted by Stem, C. (1973) Principles of Human Genetics. Freeman, San Francisco.

We can measure how well the observed frequencies agree with the expected ones with a Chi-squared test:

```{r}
# Summarizing what we have
exp_obs <- data.frame(
    `Expected freq` = expected,
    `Observed freq` = c(P, H, Q))
exp_obs

# Applying Chi-squared test
observed <- c(MM = 1787, MN = 3039, NN = 1303)
HardyWeinberg::HWChisq(observed, verbose=TRUE)
```

The $\chi^2$ statistic is very close to 0, with *P* > 0.05. Then, we conclude that the observed and expected frequencies are not different.