# Genetic constitution of a population {#intro}

## Problem 1

The following numbers of the human M-N blood groups were recorded in a sample of American Whites.

|      |      |      |
|------|------|------|
| M    | MN   | N    |
| 1787 | 3039 | 1303 |

**(1) What are the genotype frequencies observed in this sample?**

```{r}
MM <- 1787
MN <- 3039
NN <- 1303
n <- MM + MN + NN

# Genotype frequencies
P <- MM / n
P

H <- MN / n
H

Q <- NN / n
Q
```

(2) What are the gene frequencies?

Given that:

$$p = P + \frac{1}{2}H$$ and $$q = Q + \frac{1}{2}H$$

We get to:

```{r}
# Gene frequencies
p <- P + H/2
p

q <- Q + H/2
```

(3) With the gene frequencies observed, what are the genotype frequencies expected from the Hardy-Weinberg law?

Expected frequencies in the equilibrium:

$$p^2 + 2pq + q^2 = 1$$

```{r}
expected <- c(MM = p^2, 
              MN = 2*p*q, 
              NN = q^2)
expected
```

(4) How well do the observed frequencies agree with the expectation? Data from Wiener, A.S. (1943) quoted by Stem, C. (1973) Principles of Human Genetics. Freeman, San Francisco.

We can measure how well the observed frequencies agree with the expected ones with a Chi-squared test:

```{r}
# Summarizing what we have
exp_obs <- data.frame(
    `Expected freq` = expected,
    `Observed freq` = c(P, H, Q))
exp_obs

# Applying Chi-squared test
observed <- c(MM = 1787, MN = 3039, NN = 1303)
HardyWeinberg::HWChisq(observed, verbose=TRUE)
```

The $\chi^2$ statistic is very close to 0, with *P* \> 0.05. Then, we conclude that the observed and expected frequencies are not different.

## Problem 2

**About 30 per cent of people do not recognize the bitter taste of phenyl-thio-carbamate (PTC). Inability to taste it is due to a single autosomal recessive gene. What is the frequency of the non-tasting gene, assuming the population to be in Hardy-Weinberg equilibrium?**

If the trait is caused by two recessive alleles (*aa*), then the frequency of *aa* will be equal to $q^2$. As $q^2 = 0.3$, $q = \sqrt{0.3} = 0.548$.

## Problem 3

**Albinism occurs with a frequency of about 1 in 20,000 in European populations. Assuming it to be due to a single autosomal recessive gene, and assuming the population to be in Hardy-Weinberg equilibrium, what proportion of people are carriers? Only an approximate answer is needed.**

Carriers are heterozigote individuals, i.e., *2pq*. From $q^2$, we can calculate p and, hence, *2pq* as follows:

```{r}
q2 <- 1/20000 
q <- sqrt(q2)
p <- 1 - q
heterozygotes <- 2*p*q
heterozygotes
```

## Problem 4

**As an exercise in algebra, work out the gene frequency of a recessive mutant in a random-breeding population that would result in one-third of normal individuals being carriers.**

1/3 of the normal individuals as carriers means:

$$
\frac{2pq}{1-q^2} = \frac{2q(1-q)}{1-q^2} = \frac{2q(1-q)}{(1+q)(1-q)} = \frac{2q}{1+q}
$$

## Problem 5

**Three allelic variants, A, B, and C, of the red cell acid phosphatase enzyme were found in a sample of 178 English people. All genotypes were distinguishable by electrophoresis, and the frequencies in the sample were**

| Genotype | AA  | AB   | BB   | AC  | BC  | CC  |
|----------|-----|------|------|-----|-----|-----|
| Freq (%) | 9.6 | 48.3 | 34.3 | 2.8 | 5.0 | 0.0 |

**What are the gene frequencies in the sample? Why were no CC individuals found?**

As the frequency of a gene is equal to the frequency of homozygotes + 1/2 the frequency of heterozygotes, we have:

$$A = 0.096 + \frac{1}{2}(0.483 + 0.028) = 0.3515$$

$$B = 0.343 + \frac{1}{2}(0.483 + 0.05) = 0.6095$$

$$C = 0.0 + \frac{1}{2}(0.028 + 0.05) = 0.039$$ The expected frequency of CC individuals is $C^2 = 0.039^2 = 0.0015$. As we have a sample size of 178 individuals, we would expect $178 * 0.0015 = 0.27$ CC individuals (less than 1). Thus, the observed frequency (zero) is expected. We would be able to observe CC individuals if the sample size was larger.

## Problem 6

**About 7 per cent of men are colour-blind in consequence of a sex-linked recessive gene. Assuming Hardy-Weinberg equilibrium, what proportion of women are expected to be (1) carriers, and (2) colour-blind? (3) In what proportion of marriages are both husband and wife expected to be colour-blind?**

## Problem 7

**Sine oculis (so) and cinnabar (cn) are two autosomal recessive genes in Drosophila melanogaster. They are very closely linked and can be treated as if they were alleles at one locus. The 'heterozygote', so/cn, is wild-type and is distinguishable from both homozygotes; (so/so has no eyes; cn/cn has white eyes if the stock is made homozygous for another eye- colour mutant, brown, hw). In a class experiment 4 males and 4 females of an so/so stock were put in a vial together with 16 males and 16 females from a cn/cn stock and allowed to mate. There were 20 such vials. The total count of progeny, classified by genotype, was as follows.**

| so/so | so/cn | cn/cn |
|-------|-------|-------|
| 135   | 359   | 947   |


**How do these numbers differ from the Hardy-Weinberg expectations? Suggest a reason for the discrepancy.**
